Depression and suicidality as costly signals of need

### 2021-04-14 Paper.rmd
Set to word_document2. Other edits to costly signaling section.

---
title: "Depression and suicidality as costly signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is to produce signals that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever signaling is comparatively cheaper for those who have the relevant underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal than those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

@@

The current study is a refinement of a large MTurk experimental vignette pilot study (N=`r nrow(signalingdata2018b)`) that used a different vignette but very similar signals and outcomes, and a much smaller pilot study posted to reddit.com (N=XX) that used draft versions of X vignettes used here, along with the same signals and outcomes. In the MTurk pilot study, there was a single vignette in which an imagined sister claims her child needs an expensive emergency medical procedure, and asks for a very large loan. In the small reddit.com study, participants viewed and responded to two vignettes used here, each with a different signal.

@@

In the reddit.com pilot study, participants responded with lower belief and action to the second vignette compared to the first (regardless of the content or signal). To avoid such order effects, which are a problem in vignette studies that use multiple vignettes [@su2020], in the current study we decided to present each participant with a single vignette and single signal.

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. Compared to these scenarios, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner). With the "thwarted marriage" vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information). It is also possible that our research in designing the vignettes was insufficient for the population or the content of the vignette.

@@

The main limitation of this study is its lack of ecological validity, something which had to be sacrificed for any experimental design involving depression and suicide attempts. It is almost certain that participants experienced this study differently than they would if the event actually happened to them. This may have biased results in a pro-signaler direction if the lack of meaningful costs made support feel less costly or if the likely inability of participants to imagine themselves in an anonymous scenario [@hagen2006] led them to take an opportunity to signal generosity cheaply. It may have also biased participants against helping if they could not fully imagine the characters in the story as kin or important cooperative partners, while the survey's short duration may have weakened the strength of the costlier signals as bargaining tools. However, the fact that the relationships among signal cost, belief in need, and likelihood of providing support all largely fit existing theory (narrow and broad) indicates that participants paid attention to the relevant details and responded in ways that suggest some degree of external validity. For this reason, the benefits of addressing causal relationships that came through our experimental vignette approach likely outweighed the costs.

Another limitation is our uncertainty over the data quality from the Indian sample. Compared to the US sample, far more participants were filtered out by our attention checks, something which could be consistent with botting, unfamiliarity with English, or low effort responses [@kennedy2020]. Assuming this corresponded with lower-quality responses among those who passed the attention checks, the differences seen among those in the thwarted marriage vignette may simply be due to added noise rather than differences in the scenario or the effectiveness of the signals compared to those in the US.

>>>>

---
title: "Depression and suicidality as costly signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output: 
  bookdown::word_document2:
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Due to the ease with which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is to produce signals that are unprofitable to fake [@Zahavi1975]. This is possible whenever the benefits of signaling outweigh the costs for those who have the relevant underlying quality but the costs outweigh the benefits for those who lack the quality or possess it to a lesser degree [@zahavi1977; @grafen1990]. In particular, if the benefits of sending the signal are the same for signalers regardless of condition, but the signal cost outweighs this benefit for signalers without the underlying quality (a "costly signal"), then individuals with the underlying quality can evolve to send the signal and receivers can evolve to respond in mutually beneficial ways to the signal.

@@

The current study is a refinement of a large MTurk experimental vignette pilot study (N=`r nrow(signalingdata2018b)`) that used a different vignette but very similar signals and outcomes, and a much smaller pilot study posted to reddit.com (N=28) that used draft versions of 3 vignettes used here, along with the same signals and outcomes. In the MTurk pilot study, there was a single vignette in which an imagined sister claims her child needs an expensive emergency medical procedure, and asks for a very large loan. In the small reddit.com study, participants viewed and responded to two vignettes used here, each with a different signal.

@@

In the reddit.com pilot study, participants viewed two vignettes (with different signals in each one), and responded with lower belief to the second vignette compared to the first (regardless of the vignette or signal). To avoid such order effects, which are a problem in vignette studies that use multiple vignettes [@su2020], we decided to present each participant in the current study with a single vignette and single signal.

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. Compared to these scenarios, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner).

The weak signal effect in the Thwarted marriage condition, which involved the Indian sample, could indicate that our results do not generalize across cultures, undermining our adaptationist hypothesis. It could also reflect our poor understanding of contemporary Indian culture regarding dowry. The fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information).

@@

The main limitation of this study is that is has less ecological validity than real-world observations of depressed individuals interacting with their social partners. It is almost certain that participants experienced this study differently than they would if the event actually happened to them. This may have biased results in a pro-signaler direction if the lack of real costs of helping made support feel less costly or if there was a social desirability bias toward helping [@grimm2010]. It may have also biased participants against helping if they could not fully imagine the characters in the story as kin or important cooperative partners, while the survey's short duration may have weakened the strength of the costlier signals as bargaining tools.

<!-- However, the fact that the relationships among signal cost, belief in need, and likelihood of providing support all largely fit existing theory (narrow and broad) indicates that participants paid attention to the relevant details and responded in ways that suggest some degree of external validity. For this reason, the benefits of addressing causal relationships that came through our experimental vignette approach likely outweighed the costs. -->

Another limitation is that, compared to the US sample, far more Indian participants were filtered out by our attention checks, something which could be consistent with botting, unfamiliarity with English, or low effort responses [@kennedy2020]. Assuming this corresponded with lower-quality responses among those who passed the attention checks, the weak signal effect in the thwarted marriage vignette may simply be due to added noise rather than differences in the scenario or the effectiveness of the signals compared to those in the US.
